---
title: "Patient Outcome Analysis: Heart Transplant Wait-List Data"
output: github_document
---

## Executive Summary

This project analyzes the full heart transplant wait-list data set (172 patients) from the R `survival` package. We explore clinical indicators such as age, transplant status, and monitoring duration to understand patient outcomes. The goal is to demonstrate a basic Healthcare IT analytics workflow using R.

------------------------------------------------------------------------

## Tools Used

-   R
-   tidyverse
-   ggplot2
-   survival package
-   Logistic regression

------------------------------------------------------------------------

## Dataset Description

The dataset comes from the R `survival` package which contains survival data for patients on the waiting list for the Stanford heart transplant program.

For this project, we focused on these variables:

-   `Age`: patient age in years
-   `Status`: final clinical status, labeled as `Stable` or `Outcome`
-   `Monitoring_Days`: number of days observed in the dataset
-   `Transplant`: whether the patient received a transplant

------------------------------------------------------------------------

## Data Loading and Cleaning

```{r setup, message=FALSE, warning=FALSE}
# load libraries 
library(tidyverse)
library(ggplot2)

# contains professional "Heart" data set
data(heart, package = "survival")

# data cleaning
heart_clean <- heart %>%
  mutate(Age = round(age + 48), # correcting to actual age
    Status = factor(event, levels = c(0, 1), labels = c("Stable", "Outcome")), # labeling clinical outcomes
    Monitoring_Days = as.integer(stop - start),  # calculating total observation period
    Transplant = factor(transplant, levels = c(0, 1), labels = c("No", "Yes"))) %>% # labeling transplant status
  select(Age, Status, Monitoring_Days, Transplant)
heart_clean

# verify cleaned data and count
heart_count <- nrow(heart_clean) # counting 172
heart_count
```

The original `Age` variable is stored as age minus 48 in the processed dataset, we added 48 back to make the age easier to understand. We also changed the numeric outcome and transplant variables into readable labels.

------------------------------------------------------------------------

## Summary of Patient Data

```{r}
# summary table
summary_table <- heart_clean %>%
  summarize(
    Total_Patients = n(),
    Average_Age = round(mean(Age), 1),  
    Median_Age = median(Age), 
    Average_Monitoring_Days = round(mean(Monitoring_Days), 1),
    Median_Monitoring_Days = median(Monitoring_Days),
    Stable_Count = sum(Status == "Stable"),
    Outcome_Count = sum(Status == "Outcome"),
    Transplant_Count = sum(Transplant == "Yes"),
    No_Transplant_Count = sum(Transplant == "No"))
summary_table
```

This summary table gives a quick overview of the patient group. It shows the number of patients, age range, average monitoring duration, outcome count, and transplant count.

------------------------------------------------------------------------

## Exploratory Data Analysis

### Age Distribution of Heart Patients

```{r}
# creating plot 
ggplot(heart_clean, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "#3498db", color = "white") +
  labs(title = "Patient Age Distribution", x = "Age (Years)", y = "Patient Count") +
  theme_minimal()
```

The histogram shows that many patients in this dataset were middle-aged. This helps describe the patient group, but it should not be used to make broad medical conclusions because this dataset is small and specific to one heart transplant program.

------------------------------------------------------------------------

### Clinical Outcome by Transplant Status

```{r}
# outcome by transplant
outcome_by_transplant <- heart_clean %>%
  group_by(Transplant, Status) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Transplant) %>%
  mutate(Percent = round(Count / sum(Count) * 100, 1))
outcome_by_transplant

# creating plot
ggplot(outcome_by_transplant, aes(x = Transplant, y = Percent, fill = Status)) +
  geom_col(position = "dodge") +
  labs(title = "Clinical Outcome by Transplant Status",
       x = "Received Transplant", y = "Percent of Patients", fill = "Final Status") +
  theme_minimal()
```

This chart compares patient outcomes based on whether the patient received a transplant. This is useful for finding patterns in the data. However, this is observational data, so the result shows association only. It does not prove that transplant status directly caused the outcome.

------------------------------------------------------------------------

### Clinical Outcome by Age Group

```{r}
# grouping age
heart_clean <- heart_clean %>%
  mutate(
    Age_Group = case_when(
      Age < 30 ~ "Under 30",
      Age >= 30 & Age < 40 ~ "30s",
      Age >= 40 & Age < 50 ~ "40s",
      Age >= 50 & Age < 60 ~ "50s",
      Age >= 60 ~ "60+"),
    Age_Group = factor(
      Age_Group,
      levels = c("Under 30", "30s", "40s", "50s", "60+")))

# outcome clean
age_outcome <- heart_clean %>%
  group_by(Age_Group, Status) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Age_Group) %>%
  mutate(Percent = round(Count / sum(Count) * 100, 1))
age_outcome

# creating plot
ggplot(age_outcome, aes(x = Age_Group, y = Percent, fill = Status)) +
  geom_col(position = "dodge") +
  labs(title = "Clinical Outcome by Age Group",
       x = "Age Group", y = "Percent of Patients", fill = "Final Status") +
  theme_minimal()
```

This chart compares clinical outcomes across different age groups. It gives a clearer view than the age histogram alone because it connects age groups with final patient status. Some age groups may have small sample sizes, so the result should be interpreted carefully.

------------------------------------------------------------------------

### Monitoring Duration vs Clinical Status

```{r}
# creating plot 
ggplot(heart_clean, aes(x = Age, y = Monitoring_Days, color = Status)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = c("#2ecc71", "#e74c3c")) +
  labs(title = "Age vs. Monitoring Duration",
       x = "Patient Age", y = "Days of Observation",
       color = "Final Status") +
  theme_light()
```

This scatter plot shows the relationship between patient age, monitoring duration, and final clinical status. It helps identify whether outcomes are more common in shorter or longer observation periods.

------------------------------------------------------------------------

### Clinical Outcome by Monitoring Period

```{r}
# group days
heart_clean <- heart_clean %>%
  mutate(
    Monitoring_Group = case_when(
      Monitoring_Days <= 30 ~ "0-30 days",
      Monitoring_Days > 30 & Monitoring_Days <= 180 ~ "31-180 days",
      Monitoring_Days > 180 & Monitoring_Days <= 365 ~ "181-365 days",
      Monitoring_Days > 365 ~ "Over 365 days"),
    Monitoring_Group = factor(
      Monitoring_Group,
      levels = c("0-30 days", "31-180 days", "181-365 days", "Over 365 days")))

# outcome clean
monitoring_outcome <- heart_clean %>%
  group_by(Monitoring_Group, Status) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Monitoring_Group) %>%
  mutate(Percent = round(Count / sum(Count) * 100, 1))
monitoring_outcome

# creating plot
ggplot(monitoring_outcome, aes(x = Monitoring_Group, y = Percent, fill = Status)) +
  geom_col(position = "dodge") +
  labs(title = "Clinical Outcome by Monitoring Period",
       x = "Monitoring Period", y = "Percent of Patients", fill = "Final Status") +
  theme_minimal()
```

This chart compares patient outcomes by monitoring period. It gives a grouped view of the same idea shown in the scatter plot.

------------------------------------------------------------------------

## Best Predictive Model

A logistic regression model was used to estimate the probability of a clinical outcome. The predictors were age, monitoring duration, and transplant status. This model is not intended for medical decision-making. It is only used to demonstrate a basic Healthcare IT analytics workflow.

```{r}
# prepare model
heart_model <- heart_clean %>%
  mutate(
    Outcome_Binary = ifelse(Status == "Outcome", 1, 0),
    Transplant_Binary = ifelse(Transplant == "Yes", 1, 0))

# set upt model 
logistic_model <- glm(Outcome_Binary ~ Age + Monitoring_Days + Transplant_Binary,
                      data = heart_model, family = binomial)
# summary
summary(logistic_model)
```

### Model Prediction

```{r}
# prediction 
heart_model <- heart_model %>%
  mutate(
    Predicted_Probability = predict(logistic_model, type = "response"),
    Predicted_Status = ifelse(Predicted_Probability > 0.5, "Outcome", "Stable"))
head(heart_model)

# accuracy
model_accuracy <- mean(heart_model$Predicted_Status == heart_model$Status)
model_accuracy
```

The model accuracy gives a simple check of how often the predicted status matches the actual status. This is a basic model evaluation step. A stronger project could use train/test split or cross-validation, but this simple version is enough to demonstrate the workflow clearly.

------------------------------------------------------------------------

### Predicted Outcome Probability by Age

```{r}
# creating plot
ggplot(heart_model, aes(x = Age, y = Predicted_Probability, color = Status)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Predicted Probability of Clinical Outcome by Age",
    x = "Patient Age",
    y = "Predicted Probability of Outcome",
    color = "Actual Status"
  ) +
  theme_minimal()
```

This plot shows each patient's predicted probability of clinical outcome. The points are colored by actual status, so we can visually compare the model prediction with the real outcome label.

------------------------------------------------------------------------

## Key Findings

-   Many patients in this dataset were middle-aged.

-   Outcome patterns were different across transplant status groups. Age group and monitoring period comparisons made the patterns easier to see.

-   The logistic regression model showed how a simple Healthcare IT prediction workflow can be built in R.

-   The results should be interpreted carefully because this is a small observational dataset.

------------------------------------------------------------------------

## Limitations

-   The dataset is small.

-   This is observational data, so it cannot prove cause and effect.

-   Only a few variables were used in this project.

-   The model is for learning and portfolio demonstration only.

-   The results should not be used for real medical decision-making.

------------------------------------------------------------------------

## Conclusion

This project demonstrates a basic Healthcare IT analytics workflow using R. The workflow includes data cleaning, summary statistics, exploratory data analysis, grouped outcome comparison, and a simple logistic regression model. This makes the project stronger than a basic visualization-only project because it includes both descriptive analysis and basic predictive modeling.